---
title: "Hands-on Exercise 09: Modelling, Visualising and Analysing Network Data with R"
format:
  html:
    fontcolor: '#3b3b3b'
    code-line-numbers: true
author: "Moo Jia Rong"
date: "March 18, 2025"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true 
---

### **1 Overview**

In this exercise, we will model, analyse and visualise network data using R, by:

- creating graph object data frames, manipulate them using appropriate functions of dplyr, lubridate, and tidygraph,
-   building network graph visualisation using appropriate functions of ggraph,
-   computing network geometrics using tidygraph,
-   building advanced graph visualisation by incorporating the network geometrics, and
-   building interactive network visualisation using visNetwork package.

### **2 Getting STarted**

::: panel-tabset

## Loading Packages
```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts)
```

## Import Dataset
```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

## Glimpse
```{r}
glimpse(GAStech_edges)
```

:::

The output report of GAStech_edges reveals that the SentDate is treated as “Character” data type instead of date data type. This is an error! Before we continue, it is important for us to change the data type of SentDate field back to “Date”” data type.

### **3 Wrangling Data**

The code chunk below  change the data type of SentDate field back to “Date”” data type.
```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

A close examination of GAStech_edges data.frame reveals that it consists of individual e-mail flow records. This is not very useful for visualisation.

In view of this, we will aggregate the individual by date, senders, receivers, main subject and day of the week.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

### **4 Creating network objects using tidygraph**

In this section, we will create a graph data model using tidygraph package. 

#### **4.1 The tbl_graph object**

Two functions of tidygraph package can be used to create network objects, they are:

-   `tbl_graph()` creates a tbl_graph network object from nodes and edges data.

-   `as_tbl_graph()` converts network data and objects to a tbl_graph network. Below are network data and objects supported by `as_tbl_graph()`

    -   a node data.frame and an edge data.frame,
    -   data.frame, list, matrix from base,
    -   igraph from igraph,
    -   network from network,
    -   dendrogram and hclust from stats,
    -   Node from data.tree,
    -   phylo and evonet from ape, and
    -   graphNEL, graphAM, graphBAM from graph (in Bioconductor).

#### **4.2 Using tbl_graph() to build tidygraph data model**

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
GAStech_graph
```

-   The output above reveals that GAStech_graph is a tbl_graph object with 54 nodes and 4541 edges.
-   The command also prints the first six rows of “Node Data” and the first three of “Edge Data”.
-   It states that the Node Data is active. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time.

The nodes tibble data frame is activated by default, but we can change which tibble data frame is active using the `activate()` function. Thus, if we want to rearrange the rows in the edges tibble to list those with the highest “weight” first, we could use `activate()` and then `arrange()`.
```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```
### **5 Plotting Static Network Graphs with ggraph package**

#### **5.1 Plotting a basic network graph**

The code chunk below uses `ggraph()`, `geom-edge_link()` and `geom_node_point()` to plot a network graph by using GAStech_graph. 
```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

#### **5.2 Changing the default network graph theme**
In this section, we will use `theme_graph()` to remove the x and y axes. 
```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

#### **5.3 Changing the coloring of the plot**
`theme_graph()` can be used to change the coloring of the plot.

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

#### **5.4 Fruchterman and Reingold layout**
The code chunks below are used to plot the network graph using Fruchterman and Reingold layout.

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

#### **5.5 Modifying network nodes**
In this section, we will colour each node by referring to their respective departments.

#### **5.5 Modifying edges
In the code chunk below, the thickness of the edges will be mapped with the Weight variable.

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

### **6 Creating facet graphs**

#### **6.1 Working with facet_edges()**
In the code chunk below, `facet_edges()` is used.
```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```


#### **6.2 Working with facet_edges()**
The code chunk below uses `theme()` to change the position of the legend.
```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

#### **6.3 A framed facet graph**
The code chunk below adds frame to each graph.
```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

#### **6.4 Working with facet_nodes()**
In the code chunk below, `facet_nodes()` is used. Before getting started, it is advisable for you to read it’s reference guide at least once.
```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```


#### **7 Network Metrics Analysis**

#### **7.1 Computing centrality indices**
```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

#### **7.2 Visualising network metrics**

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

#### **7.3 Visualising Community**
```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))  

g + theme_graph()
```

### **8 Building Interactive Network Graph with visNetwork**

#### **8.1 Data preparation**
Before we can plot the interactive network graph, we need to prepare the data model by using the code chunk below.
```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

#### **8.2 Plotting the first interactive network graph**
The code chunk below will be used to plot an interactive network graph by using the data prepared.
```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```

#### **8.3 Working with layout**
In the code chunk below, Fruchterman and Reingold layout is used.
```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

#### **8.4 Working with visual attributes - Nodes**
`visNetwork()` looks for a field called “group” in the nodes object and colour the nodes according to the values of the group field.

The code chunk below rename Department field to group.
```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

When we rerun the code chunk below, visNetwork shades the nodes by assigning unique colour to each category in the group field.
```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

#### **8.5 Working with visual attributes - Edges**
In the code run below `visEdges()` is used to symbolise the edges.
- The argument arrows is used to define where to place the arrow.
- The smooth argument is used to plot the edges using a smooth curve.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

#### **8.6 Interactivity**
In the code chunk below, `visOptions()` is used to incorporate interactivity features in the data visualisation.

-   The argument highlightNearest highlights nearest when clicking a node.
-   The argument nodesIdSelection adds an id node selection creating an HTML select element.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### **9 Reference**
https://r4va.netlify.app/chap27