---
title: "Hands-on Exercise 08c: Analytical Mapping"
format:
  html:
    fontcolor: '#3b3b3b'
    code-line-numbers: true
author: "Moo Jia Rong"
date: "February 21, 2025"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true 
---

### **1 Overview**

In this exercise, we will use appropriate functions of tmap and tidyverse to perform the following tasks:

-   Import geospatial data in rds format into R environment.
-   Create cartographic quality choropleth maps by using appropriate tmap functions.
-   Create rate map
-   Create percentile map
-   Create boxmap

### **2 Getting Started**

In this exercise, we will be using the *NGA_wp.rds* data and the following packages: sf, tmap and tidyverse.

::: panel-tabset
## Loading Packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing Dataset

```{r}
NGA_wp <- read_rds("data/NGA_wp.rds")
```
:::

### **3 Basic Choropleth Mapping**

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of functional water point by LGAs",
            legend.outside = FALSE)

p2 <- tm_shape(NGA_wp) +
  tm_fill("total_wp",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of total  water point by LGAs",
            legend.outside = FALSE)

tmap_arrange(p2, p1, nrow = 1)
```

### **4 Choropleth Map for Rates**

#### **4.1 Deriving Proportion of Functional Water Points and Non-Functional Water Points**

We will tabulate the proportion of functional water points and the proportion of non-functional water points in each LGA. In the following code chunk, `mutate()` from dplyr package is used to derive two fields, namely pct_functional and pct_nonfunctional.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

#### **4.2 Plotting map of rate**

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_functional",
          n = 10,
          style = "equal",
          palette = "Blues",
          legend.hist = TRUE) +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Rate map of functional water point by LGAs",
            legend.outside = TRUE)
```

### **5 Extreme Value Maps**

Extreme value maps are variations of common choropleth maps where the classification is designed to highlight extreme values at the lower and upper end of the scale, with the goal of identifying outliers. These maps were developed in the spirit of spatializing EDA, i.e., adding spatial features to commonly used approaches in non-spatial EDA (Anselin 1994).

#### **5.1 Percentile Map**

The percentile map is a special type of quantile map with six specific categories: 0-1%,1-10%, 10-50%,50-90%,90-99%, and 99-100%. The corresponding breakpoints can be derived by means of the base R quantile command, passing an explicit vector of cumulative probabilities as c(0,.01,.1,.5,.9,.99,1). Note that the begin and endpoint need to be included.

##### **5.1.1 Data Preparation**

Step 1: Exclude records with NA by using the code chunk below.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2: Creating customised classification and extracting values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

##### **5.1.2 Creating the get.var function**

Firstly, we will write an R function as shown below to extract a variable (i.e. wp_nonfunctional) as a vector out of an sf data.frame.

-   arguments:
    -   vname: variable name (as character, in quotes)
    -   df: name of sf data frame
-   returns:
    -   v: vector with values (without a column name)

##### **5.1.3 Creating the get.var function**

we will write an R function as shown below to extract a variable (i.e. wp_nonfunctional) as a vector out of an sf data.frame.
```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```


##### **5.1.4 A percentile mapping function**

Next, we will write a percentile mapping function by using the code chunk below.

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}
```

##### **5.1.5 Test drive the percentile mapping function**

To run the function, type the code chunk as shown below.

```{r}
percentmap("total_wp", NGA_wp)
```

#### **5.2 Box map**

In essence, a box map is an augmented quartile map, with an additional lower and upper category. When there are lower outliers, then the starting point for the breaks is the minimum value, and the second break is the lower fence. In contrast, when there are no lower outliers, then the starting point for the breaks will be the lower fence, and the second break is the minimum value (there will be no observations that fall in the interval between the lower fence and the minimum value).

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

##### **5.2.1 Creating the boxbreaks function**

The code chunk below is an R function that creating break points for a box map.

-   arguments:
    -   v: vector with observations
    -   mult: multiplier for IQR (default 1.5)
-   returns:
    -   bb: vector with 7 break points compute quartile and fences

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

##### **5.2.2 Creating the get.var function**

The code chunk below is an R function to extract a variable as a vector out of an sf data frame.

-   arguments:
    -   vname: variable name (as character, in quotes)
    -   df: name of sf data frame
-   returns:
    -   v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

##### **5.2.3 Test drive the newly created function**

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

##### **Boxmap function**

The code chunk below is an R function to create a box map. - arguments: - vnam: variable name (as character, in quotes) - df: simple features polygon layer - legtitle: legend title - mtitle: map title - mult: multiplier for IQR - returns: - a tmap-element (plots a map)

```{r}
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

### **6 References**

[Analytical Mapping](https://r4va.netlify.app/chap23)
