---
title: "Hands-on Exercise 07: Visualising and Analysing Time-oriented Data"
format:
  html:
    fontcolor: '#3b3b3b'
    code-line-numbers: true
author: "Moo Jia Rong"
date: "February 18, 2025"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true 
---

### **1 Overview**

In this exercise, we will be:

-   plotting a calender heatmap by using ggplot2 functions,

-   plotting a cycle plot by using ggplot2 function,

-   plotting a slopegraph

-   plotting a horizon chart


### **2 Getting Started**

In this exercise, we will use the *eventlog.csv* file, which contains data on time-series cyber attack records by country, and the following packages: scales, viridis, lubridate, ggthemes, gridExtra, readxl, knitr, data.table, tidyverse

::: panel-tabset

## Loading Packages
```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes, gridExtra, readxl, knitr, data.table, tidyverse, CGPfunctions)
```


## Importing Dataset
```{r}
attacks <- read_csv("data/eventlog.csv")
```

## Examining the data
```{r}
kable(head(attacks))
```

:::

There are three columns in the dataset, namely timestamp, source_country and tz.

timestamp field stores date-time values in POSIXct format.
source_country field stores the source of the attack. It is in ISO 3166-1 alpha-2 country code.
tz field stores time zone of the source IP address.

### **3 Data Preparation**

Before we can plot the calender heatmap, two new fields namely wkday and hour need to be derived. In this step, we will write a function to perform the task.

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts, 
                        tz = tz[1], 
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
  }
```

-   `ymd_hms()` and `hour()` are from lubridate package, and
-   `weekdays()` is a base R function.

Next, we derive the attacks tibble data frame:

```{r}
wkday_levels <- c('Saturday', 'Friday', 
                  'Thursday', 'Wednesday', 
                  'Tuesday', 'Monday', 
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp, 
                   .$source_country, 
                   .$tz)) %>% 
  ungroup() %>% 
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

Table below shows the tidy tibble table after processing.
```{r}
kable(head(attacks))
```

### **4 Building Calendar Heatmaps**

```{r}
grouped <- attacks %>% 
  count(wkday, hour) %>% 
  ungroup() %>%
  na.omit()

ggplot(grouped, 
       aes(hour, 
           wkday, 
           fill = n)) + 
geom_tile(color = "white", 
          size = 0.1) + 
theme_tufte(base_family = "Helvetica") + 
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
labs(x = NULL, 
     y = NULL, 
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6) )
```

#### **4.1 Building Multiple Calendar Heatmaps**

Step 1: Deriving attack by country object

In order to identify the top 4 countries with the highest number of attacks, we need to do the following:

-   count the number of attacks by country,
-   calculate the percent of attackes by country, and
-   save the results in a tibble data frame.

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

Step 2: Prepare the tidy data frame
In this step, we extract the attack records of the top 4 countries from attacks data frame and save the data in a new tibble data frame (i.e. top4_attacks).

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

Step 3: Plotting the Multiple Calender Heatmap by using ggplot2 package.

```{r}
ggplot(top4_attacks, 
       aes(hour, 
           wkday, 
           fill = n)) + 
  geom_tile(color = "white", 
          size = 0.1) + 
  theme_tufte(base_family = "Helvetica") + 
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL, 
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6) )
```

### **5 Plotting Cycle Plot**
In this section, we will plot a cycle plot showing the time-series patterns and trend of visitor arrivals from Vietnam programmatically by using ggplot2 functions.

#### **5.1 Import Data**

For the purpose of this hands-on exercise, *arrivals_by_air.xlsx* will be used.

The code chunk below imports *arrivals_by_air.xlsx* by using `read_excel()` of readxl package and save it as a tibble data frame called air.

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

#### **5.2 Deriving month and year fields**
Next, two new fields called month and year are derived from Month-Year field.

```{r}
air$month <- factor(month(air$`Month-Year`), 
                    levels=1:12, 
                    labels=month.abb, 
                    ordered=TRUE) 
air$year <- year(ymd(air$`Month-Year`))
```

#### **5.3 Extracting the target country**

Next, the code chunk below is use to extract data for the target country (i.e. Vietnam)

```{r}
Vietnam <- air %>% 
  select(`Vietnam`, 
         month, 
         year) %>%
  filter(year >= 2010)
```


#### **5.4 Computing year average arrivals by month**

The code chunk below uses `group_by()` and `summarise()` of dplyr to compute year average arrivals by month.

```{r}
hline.data <- Vietnam %>% 
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

#### **5.5 Plotting the cycle plot**

```{r}
ggplot() + 
  geom_line(data=Vietnam,
            aes(x=year, 
                y=`Vietnam`, 
                group=month), 
            colour="black") +
  geom_hline(aes(yintercept=avgvalue), 
             data=hline.data, 
             linetype=6, 
             colour="red", 
             size=0.5) + 
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```

### **6 Plotting Slopegraph**

::: panel-tabset

## Import Data
```{r}
rice <- read_csv("data/rice.csv")
```

## Plot
```{r}
rice %>% 
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  newggslopegraph(Year, Yield, Country,
                Title = "Rice Yield of Top 11 Asian Counties",
                SubTitle = "1961-1980",
                Caption = "Prepared by: Dr. Kam Tin Seong")
```

:::

